```{r}
library(ape)
library(MVA)
library(tidyverse)
library(mclust)
library(cluster)
```

```{r}
Data <- read_csv('../Data/top2018.csv')
NumericData <- Data %>% select(4:16) %>% scale()

PrinComps <- prcomp(NumericData)
summary(PrinComps)

PCdat <- as.tibble(PrinComps$x[,1:2])

PCdat %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point()
```

# Hierarchical Clustering -------------------------------------------------

For the hierarchical clustering method clusters are based on the distance matrix. With the _single_ method, it connects the points based on the minimum distance of the clusters. However, the strange shape of the dendrogram is due to the Single Linkage resulting in "chaining", meaning the clusters just grow, instead of new clusters being formed. This can be problematic because we only need a single pair of points to be close to merge two clusters.Therefore, clusters can be too spread out and not compact enough. The _average_ method, which connects the points based on average distances of the clusters, also has the same phenomenon occurring.

```{r}
Dist <- dist(NumericData)

plot(hclust(Dist, method = 'single'))
plot(hclust(Dist, method = 'average'))
```


The _complete_ method, which connects the points based on the maximum distance of the clusters does not have any "chaining" issues. When determining what height to cut off the groups, we chose height 7 as that is the first time we start to see a lot fo tiny groups forming into larger groups.

```{r}
plot(hclust(Dist, method = "complete"))

Clusters <- cutree(hclust(Dist, method = "complete"), h = 7)
ClusterMerger <- tibble(HClusters = Clusters)
PCdat <- cbind(PCdat, ClusterMerger)

PCdat <- PCdat %>%
  mutate(HClusters = factor(HClusters))
```

When visualizing these clusters over the PC Plot it seems that cluster 3 starts to take over most of the points. When comparing it to cluster 1, it is not entirely clear why a point should be in one group compared to another. There are also a lot of little groups that are in the outskirts of the plot. While you could argue that making less clusters by increasing the height, if the clustering logic already appears to be faulty at this point it may be detrimental to the process by making too few clusters.

```{r}
PCdat %>%
  ggplot(aes(x = PC1, y = PC2, label = rownames(PCdat), color = HClusters)) +
  geom_point() +
  labs(title = "Scatterplot of Points in PC space", subtitle = "colored by cluster when cutting at a height of 7", color = "Cluster")

SilhouetteHAC <- silhouette(cutree(hclust(Dist, method = 'average'), 7), Dist)
```


# K-Means Clustering ------------------------------------------------------

K-means clustering started with choosing a number of clusters before starting. To decide how many clusters would be idea a Within-Group Sum of Squares (WGSS) plot was created using the original data. From this plot one could argue many different number of clusters would be appropriate. However, for the purposes of this analysis 15 clusters will be created. This was chosen because having too many groups would get overly confusing, especially considering there are only 100 data points. While 15 is still quite large it does get a WGSS of under about 500, more than half of where the WGSS is with only 1 group.

```{r}
n <- nrow(NumericData)
wss <- rep(0, 40)
wss[1] <- (n - 1) * sum(sapply(NumericData, var))
for (i in 2:40){
  wss[i] <- sum(kmeans(NumericData,
                       centers = i)$withinss)
}
plot(1:40, wss, type = "b", xlab = "Number of groups",
     ylab = "Within groups sum of squares",
     main = "Within-Group Sum of Squares (WGSS) Plot") # decided to use k = 15
```

The 'kmeans' function was used to create to assign each original data point to one of the 15 groups. When visualizing these clusters over the PC Plot, one can see that from a single glance these groups do not make a lot of sense for this data set. While groups tend to have clear centers and pretty consistent variability within groups, for interpreting what these results mean this method is not ideal for this data set.

```{r}
set.seed(1234)
KMeanClusts <- tibble(KClusters = kmeans(NumericData, centers = 15)$cluster)
PCdat <- bind_cols(PCdat, KMeanClusts)

PCdat %>%
  ggplot(aes(x = PC1, y = PC2, color = factor(KClusters))) +
  geom_point() +
  labs(title = "Scatterplot of Points in PC space", subtitle = "from K-Means Clustering with 15 Clusters", color = "Cluster")

SilhouetteKMeans <- silhouette(kmeans(NumericData, centers = 15)$cluster, Dist)
```

# Model Based Clustering --------------------------------------------------

For the model based clustering method assumes distributions for both the population and the sub population. While the pdfs are defined for these distributions for the purposes fo this data they will not be discussed in detail. The `Mclust` function created five separate clusters. Based on the classifications there is one point that stands out. Observation 44 is alone in its own cluster, this could either indicate that this observation is an outlier or that this clutering method is faulty.

```{r}
MClusters <- Mclust(NumericData)

MClusts <- tibble(MClusts = MClusters$classification)
PCdat <- cbind(PCdat, MClusts)
PCdat <- PCdat %>%
  mutate(MClusts = factor(MClusts))
```

When visualizing these clusters over the PC Plot, one can see that the the point in cluster 5 does appear to be out on its own compared to the other clusters. While there is some overlapping from cluster 1 across the other points, in this plot is easier to see the trends of the various clusters, especially cluster 2 and and 4.

```{r}
PCdat %>%
  ggplot(aes(x = PC1, y = PC2, color = MClusts)) +
  geom_point() +
  labs(title = "Scatterplot of Points in PC space", subtitle = "from Model-Based Clustering", color = "Cluster")

SilhouetteModelBased <- silhouette(Mclust(NumericData)$classification, Dist)
```





